---
title: "From sequences to knowledge, improving and learning from sequence alignments."
author: "Luc Blassel"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: scrbook
fontsize: 11pt
classoption:
  - twoside
csl: "./styles/nature.csl"
biblio-style: authoryear
bibliography: 
  - ./references/HPC-MSRs.bib
  - ./references/HIV-DRMs.bib
  - ./references/Sequence-Data.bib
  - ./references/Aligning-Sequences.bib
biblatexoptions: [backref=true,refsegment=chapter,defernumbers=true,sorting=none,citestyle=numeric,bibstyle=numeric,maxnames=3,maxbibnames=3,minnames=2,uniquelist=false,backend=biber]
link-citations: yes
github-repo: lucblassel/phd-manuscript
biblio-title: Global References
lof: yes
lot: yes
titlepage: tex/titlepage.tex
---

```{r, setup, include=FALSE}
library(tidyr)
library(readr)
library(dplyr)
library(ggplot2)
library(kableExtra)
library(flextable)
library(grid)
library(tiff)

options(tinytex.verbose=TRUE)
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, out.width="90%", fig.align="center", cache=TRUE)

# embed files in html output
xfun::pkg_load2(c("htmltools", "mime"))
```

\pagestyle{empty}

```{r, include=knitr::is_latex_output()}
knitr::asis_output('\n\n\\begin{comment}')
```

# Abstract {.unnumbered}

This is the Abstract

# Résumé {.unnumbered}

Ceci est le résumé

# Acknowledgments {.unnumbered}

Thanks everybody

# Glossary {.unnumbered}

What are these terms ?

```{r, include=knitr::is_latex_output()}
knitr::asis_output('\n\n\\end{comment}')
```

```{=tex}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE]{CHAPTER \thechapter} 
\fancyhead[RO]{\rightmark}
\fancyfoot[CO,CE]{\thepage}
```
# General Introduction {.unnumbered}

-   Explain shortly that 2 quite different subjects linked by alignment and sequence data.

## Organization of this manuscript {.unnumbered}

-   Organisation of the manuscript

## Publications produced {.unnumbered}

-   **Blassel, Luc**, Paul Medvedev and Rayan Chikhi. 2022. **"Mapping-friendly sequence reductions: going beyond homopolymer compression"**. *iScience*\
    [DOI goes here](https://google.com) *(Included in Chapter \@ref(HPC-paper))*

-   **Blassel, Luc**[^index-1], Anna Zhukova^1^, Christian J Villabona-Arenas, Katherine E Atkins, Stéphane Hué, and Olivier Gascuel. 2021. **"Drug Resistance Mutations in HIV: New Bioinformatics Approaches and Challenges."** *Current Opinion in Virology* 51 (December): 56--64.\
    [10.1016/j.coviro.2021.09.009](https://doi.org/10.1016/j.coviro.2021.09.009)

-   **Blassel, Luc**, Anna Tostevin, Christian Julian Villabona-Arenas, MartinePeeters, Stéphane Hué, and Olivier Gascuel. 2021. **"Using Machine Learning and Big Data to Explore the Drug Resistance Landscape in HIV."** *PLOS Computational Biology* 17 (8): e1008873.\
    [10.1371/journal.pcbi.1008873](https://doi.org/10.1371/journal.pcbi.1008873). *(Included in Chapter \@ref(HIV-paper))*

-   Zhukova, Anna, **Luc Blassel**, Frédéric Lemoine, Marie Morel, JakubVoznica, and Olivier Gascuel. 2021. **"Origin, Evolution and Global Spread of SARS-CoV-2."** *Comptes Rendus. Biologies* 344 (1): 57--75.\
    [10.5802/crbiol.29](https://doi.org/10.5802/crbiol.29).

-   Lemoine, Frédéric, **Luc Blassel**, Jakub Voznica, and Olivier Gascuel.2020. **"COVID-Align: accurate online alignment of hCoV-19 genomes using a profile HMM"** *Bioinformatics*, 37 (12): 1761-1762.\
    [10.1093/bioinformatics/btaa871](https://doi.org/10.1093/bioinformatics/btaa871).


[^index-1]: co-first authors: Luc Blassel and Anna Zhukova
