First, I would like to thank all the members of my Jury for accepting to evaluate the work that I have done over the past few years. I would like to thank more specifically Macha Nikolski and Brona Brejova for reading and reviewing my manuscript as well as providing helpful feedback on my work. Second, I would like to thank the people that supervised me and helped me along the way. Thank you to Paul for your valuable insights. Thank you to Rayan for supervising me in these last years of this project, and thank you Olivier for doing so in the first few years. I would also like to thank Didier for helping me navigate the turmoil. 

For most of my PhD (except for a few months in early 2020…), I spent my time on the 6th  floor of the Omics building of the Institut Pasteur, in the \textit{Evolutionary Bioinformatics} lab. Thank you to all my office mates for making my time on the 6th floor so pleasant. Thank you Anna, Fred and Jakub for all the discussions and coffee breaks. A special thanks to my good-colleague Marie, thanks to you I now have an ingrained response of looking up and to the right whenever I hear an office chair move. Thank you also to the 6th floor adoptees, Jerome, Andrew, as well as Arthur, the earliest of them. I would also like to thank my labmates from the \textit{Sequence Bioinformatics} lab: Camilla, Francesco, Yoann, Riccardo, Luca and Téo. 

Of course, as studious as I might wish to be, I did spend 4 years at the office. I would therefore like to thank all my friends that have made these enjoyable. Thank you Grimault, Nathalie, Théo for the parties and the German escapades. Thank you as well Eva, Max and Flore for everything. It has been a while since Balzac, but I am very glad we still all hang out! Thank you as well to Dylan and Joshua, I started this adventure with you as roomates and enjoyed it very much. Finally thank you to Marie and Jean for many, many, celebrations!

Next up on the thank you list is, of course, my family. I would like to thank my mom, dad and brothers, Noé and Tom, for supporting me along my journey. Thank you as well to my grandparents: Mamette, André and Grandma for believing in me and helping me along the way. To all the others from Tokyo to Santa Cruz, it warms my heart to know I have a worldwide support network, thank you!

Finally, Fatima. I am so glad I met you during this adventure. Thank you for supporting me, talking things through with me and helping me along the way. I am very grateful that you are part of my life, and I feel better looking to the future knowing you will be by my side.

As parting words, I would like to dedicate this work to Pierre Blassel and John F. Murray. They made me, in part, what I am today and were essential in instilling in me the curiosity and scientific interest that are so necessary to pursue this type of endeavor. Although they could not see me finish, I would like to thank them from the bottom of my heart and I can only hope that they would be proud of me.
